A Short and Mechanized Consistency Proof for Dependent Type Theory

### 2024-02-15 paper/paper.tex
Add the coq pi rule

In Section~\ref{sec:logrelmech}, we discuss the details about our
mechanization, including our use of existing libraries such as
Autosubst for handling bindings and CoqHammer for general-purpose
automation.
In Section~\ref{sec:relatedwork}, we discuss how our proof
relates to existing proofs by logical relations for dependent types in
the literature.

@@

In Coq, there is a distinction between computable functions and
relations that can later be proven to be functional. The former can be
viewed as a strict subset of the latter in axiom-free Coq. To be more
precise, given a relation \texttt{R : A -> B -> Prop} subject to the
totality and functionality constraints ($\forall$ \texttt{a} $\in$
\texttt{A}, there exists a unique \texttt{b} $\in$ \texttt{B} such
that \texttt{R a b} is inhabited), we do not immediately obtain a function
\texttt{F : A -> B} such that $\forall$ \texttt{a} $\in$ \texttt{A},
\texttt{R a (F a)} is inhabited. However, the functional side
conditions of a relation is clunky to express and tend to block
automation. A simple workaround is to assume the choice axiom, which
is known to be consistent and allows us to induce a function \texttt{F
: A -> B} once we have shown the relation \texttt{R : A -> B -> Prop}
is functional. This approach would make our Coq development match
the text version of our proof from Section~\ref{sec:logrelproof} more
closely.

However, we choose instead an axiom-free workaround and define
\rref{T-Pi} as follows in our Coq mechanization.

Instead of constraining \texttt{F} to be a function or a functional
relation, we leave \texttt{F} as an unrestricted relation that is
total on input from the set $[[S]]$ and ...TODO...

This formulation allows us to derive Lemma~\ref{lemma:piintroalt}
before we even show that our logical relation is
functional/deterministic, but does not affect the proof structure
otherwise. We choose to keep the discrepancy since the skolemization
process is typically expressed in terms of function symbols rather
than relation symbols.

\subsection{Automation}
\label{sec:automation}
Our Coq mechanization heavily uses automation, though instead of
defining custom tactics, we rely mostly on off-the-shelf tools such as
Autosubst 2 and CoqHammer.

@@

For other automation tasks that are not specific to binding, we use
the powerful \texttt{sauto} tactic provided by CoqHammer to write
short and declarative proofs. For example, here is a one-line proof of
the triangle property about parallel reduction, from which the diamond
property (Lemma~\ref{lemma:pardiamond}) follows as a corollary.
\begin{minted}{coq}
Lemma par_triangle a : forall b, Par a b ->  Par b (tstar a).
Proof.
  apply tstar_ind; hauto lq:on inv:Par use:Par_refl,par_cong ctrs:Par.
Qed.
\end{minted}
In prose, the triangle property can be proven by induction over the
graph of the \texttt{tstar} function, which stands for the Takahashi
translation~\citep{takahashi-parallel-reduction}. Options \texttt{inv:Par} and
\texttt{ctrs:Par} says that the proof involves inverting and
constructing of the derivations of parallel
reduction. The option \texttt{use:Par\_refl,par\_cong} allows the
automation tactic to use
the reflexivity and congruence properties of parallel
reduction as lemmas.

>>>>

In Section~\ref{sec:logrelmech}, we discuss the details about our
mechanization, including our use of existing libraries such as
Autosubst 2~\citep{autosubst2} for handling bindings and CoqHammer~\citep{czajka2018hammer} for general-purpose
automation.
In Section~\ref{sec:relatedwork}, we discuss how our proof
relates to existing proofs by logical relations for dependent types in
the literature.

@@

In Coq, there is a distinction between computable functions and
relations that can later be proven to be functional. The former can be
viewed as a strict subset of the latter in axiom-free Coq. To be more
precise, given a relation \texttt{R : A -> B -> Prop} subject to the
totality and functionality constraints ($\forall$ \texttt{a} $\in$
\texttt{A}, there exists a unique \texttt{b} $\in$ \texttt{B} such
that \texttt{R a b} is inhabited), we do not immediately obtain a function
\texttt{F : A -> B} such that $\forall$ \texttt{a} $\in$ \texttt{A},
\texttt{R a (F a)} is inhabited. However, the functional side
conditions of a relation is clunky to express and tend to block
automation. A simple workaround is to assume the axiom of unique choice, which
is known to be consistent with Coq and allows us to induce a function \texttt{F
: A -> B} once we have shown the relation \texttt{R : A -> B -> Prop}
is functional. This approach would make our Coq development match
the text version of our proof from Section~\ref{sec:logrelproof} more
closely.

However, we choose instead an axiom-free workaround and define
\rref{I-Pi} as follows in our Coq mechanization.
\begin{center}
  \drule[]{I-PiCoq}
\end{center}
It should be easy to verify that the preconditions of
\rref{I-Pi}, \rref{I-PiAlt}, and \rref{I-PiCoq} are all
equivalent. After establishing Lemma~{lemma:logreldeter}, it is
possible to further show that the conclusions of the rules are
equivalent, too.

% This formulation allows us to derive Lemma~\ref{lemma:piintroalt}
% before we even show that our logical relation is
% functional/deterministic, but does not affect the proof structure
% otherwise.
We choose to keep this discrepancy between the Coq development and the
logical relation presented in Section~\ref{sec:logreldep} since the
skolemization process is more intuitively expressed in terms of
function symbols rather than relation symbols. Otherwise, we do not
see a clear advantage of \rref{I-PiCoq} over \rref{I-Pi} in set
theory, where there is no distinction between computable functions and
functions in general.

\subsection{Automation}
\label{sec:automation}
Our Coq mechanization heavily uses automation, though instead of
defining custom tactics, we rely mostly on off-the-shelf tools such as
Autosubst 2~\citep{autosubst2} and CoqHammer~\citep{czajka2018hammer}.

@@

For other automation tasks that are not specific to binding, we use
the powerful \texttt{sauto} tactic provided by CoqHammer to write
short and declarative proofs. For example, here is a one-line proof of
the triangle property about parallel reduction, from which the diamond
property (Lemma~\ref{lemma:pardiamond}) follows as a corollary.
\begin{minted}{coq}
Lemma par_triangle a : forall b, Par a b ->  Par b (tstar a).
Proof.
  apply tstar_ind; hauto lq:on inv:Par use:Par_refl,par_cong ctrs:Par.
Qed.
\end{minted}
In prose, the triangle property can be proven by induction over the
graph of the \texttt{tstar} function, which stands for the Takahashi
translation~\citep{takahashi-parallel-reduction}. Options \texttt{inv:Par} and
\texttt{ctrs:Par} say that the proof involves inverting and
constructing of the derivations of parallel
reduction. The option \texttt{use:Par\_refl,par\_cong} allows the
automation tactic to use
the reflexivity and congruence properties of parallel
reduction as lemmas.
